Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1) \note{\scriptsize Authors, please add/check middle initials and affiliations}

### 2014-04-22 WSSSPE1_report/WSSSPE1_summary_report.tex
adding Berriman affiliation
Career paths and definition issues as refs to cross-cutting.

The panel on developing and supporting software examined the challenges
around scientific software development and support, mainly focused on research
groups that also produce code in various forms. There was widespread agreement
that developing and maintaining software is hard, but best practices can help.
Several participants added that documentation is not just for users, and writing
application programming interface (API) documentation, tutorials for building and deploying software, together with
documented development practices can be very helpful in bringing  new developers
into a project efficiently.

@@

\subsection{Communities and Career Paths}

Communities are extremely important in software projects, and both their
building and continued engagement need attention during the project life cycle.
Several of the submitted papers discussed how communities have been built around
projects, and what is needed to enable a project to grow. Among these are
public source-code hosting, mailing lists, documentation, wikis, bug trackers, software
downloads, continuous integration, software quality dashboards, and of course,
a general web presence to tie all of these things together.

There was extended discussion about users not answering the questions of other
users. Several participants
offered counterexamples, such as mailing lists where developers do not participate
as much due to users being more active, or whether the ``core
team'' can end up setting unrealistic expectations by doing too much. Team Geek~\cite{opac-b1134063}
and Turk's paper on scaling code in the human dimension~\cite{Turk:2013:SCH:2484762.2484782}  discuss how development
lists that are welcoming to people actually have many more people contributing.

Recruiting and retaining personnel in this area is hard,
with one of the major reasons being no long-term career paths (especially when
compared with industry). This is further discussed in~\S\ref{sec:career-tracks}.

%moved to cross-cutting issues
%
%How should software development be balanced with
%research? It is apparent that things are beginning to improve, such as
%incentives for software development in the form of altmetrics, tenure committee
%consideration, and NSF ``products'' vs ``publications.'' It was noted it can be very
%difficult to measure where people are using small bits of your code.

\subsection{Defining Sustainability for Scientific Software}

The question of what ``sustainable'' means in the context of software was
raised (see \S\ref{sec:defining-sustainability}.) The resources at OSS Watch~\cite{osswatch-web}   discuss what
to look for when choosing software for procurement, or reuse in further software
development. Regardless of the license and development model that will be used,
the future of the project must be considered. Even if a particular piece of
software satisfies today's requirements, will it continue to do so in the
future? Important questions include whether the supplier will still be around in five
years' time, will it still care for its customers' needs, will it be responsive to
bug reports and feature requests? Should you tie your investment to a single
supplier using a proprietary product, or ensure the project uses an OSI-approved
license, and can outlive any single entity if the software is still important.

What is the overarching goal of sustainability? Is it reproducible science,
persistence, quality, something else? How should success be measured in this
context? Is there some metric that can be used to determine when you have
achieved sustainable software, or is this an ongoing process with no clear
endpoint. For truly sustainable software there should be no endpoint, as the
software products continue to be used and useful beyond the single institution,
grant, and developer/development team. Sustainability must be addressed
throughout the project life cycle.

>>>>

The panel on Developing and Supporting Software examined the challenges
around scientific software development and support, mainly focused on research
groups that in addition to pursuing research also produce code in various forms. There was widespread agreement
that developing and maintaining software is hard, but best practices can help.
Several participants added that documentation is not just for users, and writing
application programming interface (API) documentation, tutorials for building and deploying software, together with
documented development practices can be very helpful in bringing  new developers
into a project.

Two subjects that prominently surfaced in this panel also came up
throughout other parts of the workshop, and are therefore deferred to
the section on Cross-cutting Issues (\S\ref{sec:cross-cutting}). These
are the lack of long-term career paths for specialists in the various
software development and support areas (see
\S\ref{sec:career-tracks}), and the question of what ``sustainable''
should mean in the context of software (see
\S\ref{sec:defining-sustainability}).

@@

\subsection{The Importance of Communities}

Communities are extremely important in software projects, and both their
building and continued engagement need attention during the project life cycle.
Several of the submitted papers \todo{which ones? should cite here.} discussed how communities have been built around
projects, and what is needed to enable a project to grow. The latter
includes 
public source-code hosting, mailing lists, documentation, wikis, bug trackers, software
downloads, continuous integration, software quality dashboards, and of course,
a general web presence to tie a project's channels and artifacts together.

There was extended discussion about the challenge of fostering
communities in which users help each other, rather than always
deferring to the developers of project to answer user queries. Participants
offered examples that this is indeed possible, such as mailing lists
in which developers do not participate
much because users actively respond to questions from other users, but
also asked whether by doing too much the ``core
team'' could end up setting unrealistic expectations. Team Geek~\cite{opac-b1134063}
and Turk's paper on scaling code in the human dimension~\cite{Turk:2013:SCH:2484762.2484782}  discuss how development
lists tend to  have many more people contributing when they are welcoming to people.
